---
title: "Getting Started with mimicsurv"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mimicsurv}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mimicsurv)
```

## Introduction

The `mimicsurv` package provides tools for survival analysis from published Kaplan-Meier tables using the person-years method. This vignette demonstrates the basic usage of the package functions.

## Basic Survival Analysis

### Extracting Results from Kaplan-Meier Tables

The main function `extractfromKM()` estimates hazard rates and median survival time from published Kaplan-Meier data:

```{r basic-analysis}
# Example data from a hypothetical study
time_points <- c(0, 6, 12, 18, 24)
n_risk <- c(200, 150, 100, 60, 35)
n_censored <- c(0, 10, 20, 30, 40)

# Extract survival analysis results
result <- extractfromKM(time_points, n_risk, n_censored)

# View hazard table
print(result$hazard_table)

# View median survival time
cat("Median survival time:", result$median_survival, "months\n")
```

### Understanding the Results

The hazard table shows:

- **interval**: Time intervals for hazard estimation
- **n_at_risk_start**: Number of patients at risk at the start of each interval
- **n_censored_interval**: Number of patients censored within each interval
- **n_events**: Number of events (estimated) within each interval
- **hazard_rate**: Estimated hazard rate (events per person-time)

## Mathematical Background

The package implements the person-years method with the following key concepts:

### Person-Time Calculation

Person-time is calculated using the trapezoidal rule:

$$\text{Person-time} = \frac{n_{\text{risk},i} + n_{\text{risk},i+1}}{2} \times (t_{i+1} - t_i)$$

### Hazard Rate Estimation

Hazard rates are estimated as:

$$\lambda_i = \frac{\text{Number of events}_i}{\text{Person-time}_i}$$

### Median Survival Time

For piecewise exponential survival, the median is found where:

$$S(t) = \exp\left(-\sum_{j=1}^{k} \lambda_j \Delta t_j\right) = 0.5$$

## Simulation and Validation

### Generating Simulated Data

```{r simulation}
# Define piecewise exponential parameters
time_points <- c(0, 6, 12, 18, 24)
hazard_rates <- c(0.05, 0.08, 0.12, 0.15)

# Simulate survival data
sim_data <- simPE(
  n = 100,  # Reduced sample size for faster execution
  time_points = time_points,
  hazard_rates = hazard_rates,
  max_time = 24,
  censoring_prob = 0.1  # Reduced censoring probability
)

# Check event rate
table(sim_data$status)
```

### Creating Summary Tables

```{r summary-table}
# Create Kaplan-Meier style summary
km_summary <- summaryKM(sim_data, time_points)

print("Number at risk:")
print(km_summary$n_risk)

print("Cumulative censored:")
print(km_summary$n_censored_cumulative)
```

## Validation Study

### Comprehensive Validation

The `validate_mimicsurv()` function provides comprehensive validation of the hazard estimation methods:

```{r validation, eval=FALSE}
# Run validation study (small example for demonstration)
validation_result <- validate_mimicsurv(
  true_time_points = time_points,
  true_hazard_rates = hazard_rates,
  sample_sizes = c(100, 200),
  n_simulations = 50,
  censoring_prob = 0.15
)

# View summary statistics
print(validation_result$summary_statistics)

# View detailed results for one sample size
print(validation_result$detailed_results$n_200$hazard_results)
```

### Method Validation Example

Here's a practical example showing how well our method recovers known parameters:

```{r practical-validation}
# Set known "true" parameters
true_times <- c(0, 6, 12, 18, 24)
true_hazards <- c(0.04, 0.055, 0.06, 0.05)  # Monthly hazard rates

# Calculate true median survival
true_median <- getMediansurv(true_times, true_hazards)
cat("True median survival:", round(true_median, 1), "months\n")

# Simulate data using these parameters
set.seed(123)  # For reproducible results
sim_data <- simPE(
  n = 500,  # Larger sample for better precision
  time_points = true_times,
  hazard_rates = true_hazards,
  max_time = 24,
  censoring_prob = 0.1
)

# Create KM table from simulated data
km_table <- summaryKM(sim_data, true_times)

# Estimate parameters from the KM table
estimated_result <- extractfromKM(
  time_points = true_times,
  n_risk = km_table$n_risk,
  n_censored = km_table$n_censored_cumulative
)

# Compare results
comparison <- data.frame(
  Interval = estimated_result$hazard_table$interval,
  True_Hazard = true_hazards,
  Estimated_Hazard = estimated_result$hazard_table$hazard_rate,
  Relative_Error = (estimated_result$hazard_table$hazard_rate - true_hazards) / true_hazards * 100
)

print("Validation Results:")
print(comparison)

cat("\nMedian Survival Comparison:\n")
cat("True median:", round(true_median, 1), "months\n")
cat("Estimated median:", round(estimated_result$median_survival, 1), "months\n")
if (!is.na(estimated_result$median_survival)) {
  cat("Relative error:", round((estimated_result$median_survival - true_median)/true_median * 100, 1), "%\n")
}
```

## Advanced Usage

### Calculating True Median Survival

You can calculate the theoretical median survival time directly:

```{r true-median}
# Calculate true median for comparison
true_median <- getMediansurv(time_points, hazard_rates)
cat("True median survival:", round(true_median, 2), "months\n")
```

### Handling Different Scenarios

The package handles various scenarios:

- **High censoring rates**: Appropriate for studies with substantial loss to follow-up
- **Multiple intervals**: Flexible number of time intervals
- **Missing median**: Returns NA when median survival is not reached

## Best Practices

1. **Data Quality**: Ensure input data is consistent (number at risk should decrease over time)
2. **Interval Selection**: Use meaningful time intervals that reflect clinical practice
3. **Validation**: Use simulation studies to validate results when possible
4. **Interpretation**: Consider the assumptions of piecewise exponential survival

## Conclusion

The `mimicsurv` package provides a comprehensive framework for extracting survival information from published Kaplan-Meier data. The combination of estimation and validation functions makes it a valuable tool for meta-analyses and secondary data analyses in survival research.
